Governing through patronage: the bargain for education in decentralized Brazil

### 2020-09-11 paper/paper.tex
redraft paper

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors ultimately cause instability in the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration. Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

@@

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015, gulzar_politicians_2017}. Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services \citep{ferraz_corrupting_2012}. It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy \citep{kohli_state-directed_2004, evans_embedded_1995}.

>>>>

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can these educational outcomes be improved? Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy experiments.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo? Who holds power over public education, and whose interests is it meant to serve?

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors ultimately cause turnover in the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage. Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

@@

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I also bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.
